Projit: An Open Source tool for Decoupled Data Science

### 2023-10-12 docs/paper/paper.tex
Paper update

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

Progress in scientific research depends heavily on accurate record keeping of previous
experimental approaches and results. Failure to maintain these records impedes progress
by making it difficult to reproduce work, or imposing the costs of repeatedly testing 
failed lines of experimentation. This cost becomes particularly high in the age of the
reproducibility crisis where many teams are running the same experiments in parallel, without
knowledge of each others work, and producing a factory line of un-reproducible results 
\cite{Ioannidis2005}.

Many Machine Learning experimentation frameworks focus on the task of making experiments
easier to execute and deploy into production systems\cite{Alberti:2018,MolnerDomenech:2020}. 
To do so they are often constructed for specific combinations of technology. 
This is often necessary limitation to either make a project feasible or enable efficiency, 
but it has the side-effect of limiting general applicability.
Many other eScience frameworks focus on the lineage and management of data, referred to as 
the so-called provenance problem \cite{Sahoo:2008,Conquest:2021}.
The goal of the provenance frameworks is sufficient auditibiliy of data that will 
render eScience transparent and repeatable. This can be auditing of data from multiple 
source systems, or auditing of logs generated during data processing\cite{Ferdous2020}. 
At the extreme we can seek to quantify every transformation that happens to data in 
the course of processing\cite{Sahoo2009}. Regardless of the
specific data to be audited, these frameworks focus on developing unified systems and 
processes so that auditing can be easily performed over many projects.

In addition to systems for storage of data, eScience application may include facilities
for orchestration of data processes and external services\cite{Subramanian2013}, 
requests for experiments with specific parameters\cite{Hunter:2005}, 
or integrated analysis of results, generation of insights and documentation. 
Other frameworks and approaches in eScience focus on understanding how to do large 
scale collaborative science, or facilitate meta-level learning of various 
kinds\cite{Hunter:2005,Liu:2023}. The better we track the
process of science as a whole, the better we can understand both how to improve 
scientific processes as well as data mine the history of experimental results for 
phenomena that are difficult to detect.

@@

In this work we argue for development of data science frameworks that impose minimal 
requirements, both in terms of appplication domains and underlying technologies. 
We present a design framework for building decoupled data science tools that can 
improve efficiency and replication through standardisation, without
unreasonable impositions on design decisions. We describe the design of a open 
source project integration tool (\textit{projit}) that can be used either as a 
Command Line Application (CLI) or python API. Internally \textit{projit} depends 
only on a metadata store that uses the general purpose JSON format. 
As such it is trivial for developers to build interfaces in other
languages, or devise web service APIs for decentralised versions. We explore a 
case study of comparing results across multiple projects for which we have used 
the \textit{projit} application to manage our metadata.

@@

We begin by discussing all desirable elements required of an open science framework. 
These are drawn from observations of both how collaborative science works and the 
successful components of distributed scientific endeavours. These requirements are 
drawn from both sciences that are typically dependent on computational frameworks 
(computer science, bioinformatics, physics) and those that generally are not
(social science, psychology).

@@

@@

@@

We have utilised the projit application across multiple data science projects to store 
reusable datasets and the results of all experiments. Additionally, the metadata store 
contains information about the number of times each experiment has been executed, 
and the execution time utilised on each run. This allows us to generate an ad hoc 
script that can compare projects in terms of the data used, the number of experiments
conducted and the total execution time. This script is constructed for illustrative 
purposes to show that the projit tool can permit arbitrary meta-analysis of projects 
through the standardised metadata stored across git repositories.

@@

>>>>

\usepackage{listings}
\usepackage{graphicx}
\usepackage{graphics}
\usepackage{geometry}
\usepackage{xcolor}
\usepackage{anyfontsize}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}
\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=b,
    keepspaces=true,
    numbers=left,
    numbersep=5pt,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=2
}
\lstset{style=mystyle}

@@

Progress in scientific research depends heavily on accurate record keeping of previous
experimental approaches and results. Failure to maintain these records impedes progress
by making it difficult to reproduce work, or imposing the costs of repeatedly testing 
failed lines of experimentation. This cost becomes particularly high in the age of the
reproducibility crisis where many teams are running the same experiments in parallel, without
knowledge of each others work, and producing a factory line of un-reproducible results 
\cite{Ioannidis2005}. This cost is compounded by the problems of increasingly 'Big Data' 
based science, where the logistics of maintaining and tracking data sets is challenging.
While big data may offer many opportunities cross-disciplinary and algorithmic science\cite{Schmitt2015}, 
there is a strong concern that this will be limited\cite{Succi2019}, 
and that the core scientific tasks of conceiving,
designing and running experiments needs to be managed at an ever increasing scale.

Platforms for eScience offer a variety of solutions for these problems including  
tracking the lineage and management of data, referred to as 
the provenance problem \cite{Sahoo:2008,Conquest:2021}.
The goal of provenance frameworks is sufficient auditibiliy of data that will 
render eScience transparent and repeatable. This can be auditing of data from multiple 
source systems, or auditing of logs generated during data processing\cite{Ferdous2020}. 
At the extreme we can seek to quantify every transformation that happens to data in 
the course of processing\cite{Sahoo2009}. Regardless of the
specific data to be audited, these frameworks focus on developing unified systems and 
processes so that auditing can be easily performed over many projects. It has been 
argued that tracking meta-data is critical for the endeavour of large scale open and auditable 
science\cite{Reznik2022}. It is only through meta-data that we are able to build frameworks
that can understand and assist the scientific process.

In addition to systems for tracking data, eScience application may include facilities
for orchestration of data processes and external services\cite{Subramanian2013}, 
requests for experiments with specific parameters\cite{Hunter:2005}, 
or integrated analysis of results, generation of insights and documentation. 
Other frameworks and approaches in eScience focus on understanding how to do large 
scale collaborative science, or facilitate meta-level learning of various 
kinds\cite{Hunter:2005,Liu:2023}. The better we track the
process of science as a whole, the better we can understand both how to improve 
scientific processes as well as data mine the history of experimental results for 
phenomena that are difficult to detect.

@@

Many Machine Learning experimentation frameworks focus on the task of making experiments
easier to execute and deploy into production systems\cite{Alberti:2018,MolnerDomenech:2020}.
To do so they are often constructed for specific combinations of technology.
This is often necessary limitation to either make a project feasible or enable efficiency,
but it has the side-effect of limiting general applicability. An ideal data science
framework would allow open ended experimentation, with an ability for easy tracking
and comparison of these experiments.

In this work we argue for the development of data science frameworks that impose minimal 
requirements, both in terms of appplication domains and underlying technologies. 
We present a design framework for building decoupled data science tools that can 
improve efficiency and replication through standardisation, without
unreasonable impositions on design decisions. We describe the design of a open 
source meta-data tracking project integration tool (\textit{projit}) that can be used either as a 
Command Line Application (CLI) or python API. Internally \textit{projit} depends 
only on a simple metadata store that uses the general purpose JSON format. 
As such it is trivial for developers to build interfaces in other
languages, or devise web service APIs for decentralised versions. We explore a 
case study for analysing a project for which we have used 
the \textit{projit} application to manage our experiment metadata.

@@

We begin by discussing the desirable features of an open science framework. 
These are drawn from observations of both how collaborative science works and the 
successful components of distributed scientific endeavours. These requirements are 
drawn from both sciences that are typically dependent on computational frameworks 
(computer science, bioinformatics, physics) and those that generally are not
(social science, psychology).

@@

\begin{lstlisting}[language=BASH,label={code:install}, caption=Installation and Invocation of Projit CLI]
pip install projit

projit init "Test Project"
\end{lstlisting}

@@

\begin{lstlisting}[language=Python,label={code:usage}, caption=Usage of Projit Python Library]
import projit as pit

experiment_name = "My Experiment"
project = pit.projit_load()
exec_id = project.start_experiment(experiment_name, sys.argv[0], params={})
#
# EXPERIMENT EXECUTION CODE 
#
project.end_experiment(experiment_name, exec_id)
\end{lstlisting}

@@

We have utilised the projit application across multiple data science projects to store 
references to reusable datasets and experimental resultss. Additionally, the metadata store 
contains information about the number of times each experiment has been executed, 
and the execution time utilised on each run. This allows us to conduct an after-the-fact
investigation into these projects. In this section we demonstrate how some details of these
projects can be revealed through the project meta-data.

\subsection{Systematic Review with Machine Learning}

In a recent paper we explored the use of machine learning algorithms for filtering a initial
set of article abstracts as part of a systematic review\cite{HawkinsTivey2023}. 
The resulting paper focused on four
different machine learning strategies, the first of which were standard machine learning techniques
trained on simple features, such as the age of the article, the number of authors, the number of 
matching keywords in the abstract etc. The presentation of results in this paper created the impression
that all of these baseline models were executed in the first round of experimentation. By querying
the \emph{projit} meta-data store to fine the first execution time of each experiment, we created
Table \ref{tab:execs}, which shows that he Naive Bayes and SVM models were in fact executed last.

\begin{table}
\caption{Experiment Executions}
\label{tab:execs}
\resizebox{\columnwidth}{!}{%
\begin{tabular}{|l|r|r|}
\toprule
Experiment            &Execution Count    &First Execution     \\
\midrule
Dataset Size         &1                   &2   \\
Variables            &11                  &1      \\
Outlier Proportion   &2                   &0  \\
Std Multiplier       &6                   &     \\
\bottomrule
\end{tabular}
}
\end{table}

In our case this was because extended literature reviews indicated that many people working on this
problem previously tended to rely on one of these two techniques. It was decided that we should include
them in our baselines, as a reasonable reviewer response would be to ask why we did not include these
models for comparison (given their prevalance in the literature). An ability to query meta-data about
experimentation, can reveal clues as to the way a piece of research was conducted, and raise valuable
questions that might be needed to detect suspicious activity.

@@

Modern science has faced multiple problems with repeatability, credibility and feasibility. The problems
we want to solve require more and more large scale digital assets and technologies, and we require methods 
for tracking and managing this growing complexity. These problems exist in academic science and in industry
practices like data science, where organisations are investing to create reusable technologies and 
competitive advantages.

We have argued for a decoupled approach to building eScience tools. One that draws inspiration from the Unix
philosophy of make simple interoperable tools. This approach will allow us to create an open and easily 
adaptable set of tools that can support many kinds of scientific projects.

We have implemented a simple meta-data management tool using these ideas and released it as an open source 
library and python package. The \emph{projit} tool can be used to track datasets, experiments and results
for any project that uses algorithms or scripting.

Finally, we provided a simple example of using this meta-data to query the experimental details behind one
of our previous research projects. Highlighting how \emph{projit} can be used to audit a project and identify
the experimental realities behind a set of results.
